\documentclass[10pt]{article}

\usepackage{hicss51}
\usepackage{times}
\usepackage[none]{hyphenat}
\usepackage{url}
\usepackage{latexsym}
\usepackage{indentfirst}
\usepackage{graphicx}

\newcommand{\sansserifformat}[1]{\fontfamily{cmss}{ #1}}%

\let\proof\relax\let\endproof\relax\usepackage{amsthm}
\let\labelindent\relax\usepackage{enumitem}% NEEDED TO MAKE HYPERREF LIKE REFERENCES
\makeatletter
\let\NAT@parse\undefined
\makeatother
\usepackage{ragged2e}
\usepackage[bookmarks=false,colorlinks,urlcolor=blue,%
linkcolor=magenta,citecolor=red,linktocpage=true,breaklinks=true]{hyperref}
\usepackage{xspace}
\usepackage{graphics}
\usepackage{inconsolata}
\usepackage{cleveref}
\usepackage{xcolor}
\usepackage{fancyvrb}
\usepackage{listings}
\usepackage{xparse}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{pgf-umlsd}
\usepackage{microtype}

\usepackage[backend=bibtex,style=ieee,firstinits,maxnames=2,minnames=1]{biblatex}
\addbibresource{biblio.bib}
\setlength{\biblabelsep}{.5em}
\setlength{\bibitemsep}{1pt}

\include{macros}

\usepackage[textsize=tiny,linecolor=teal,%
bordercolor=teal,backgroundcolor=lime!25]{todonotes} % for margin
\setlength{\marginparwidth}{2cm}
  
\title{A Language-based Approach for Interoperability of IoT
Platforms}

\author{Maurizio Gabbrielli, Saverio Giallorenzo, Ivan Lanese, and Stefano Pio
Zingaro 
\\
Universit\`a di Bologna / INRIA \\
{\underline{maurizio.gabbrielli@unibo.it}, 
  \underline{ saverio.giallorenzo@gmail.com}}\\
  {\underline{ivan.lanese@gmail.com},
  \underline{stefanopio.zingaro@unibo.it}
  } \\
}

\date{}

\begin{document}

\maketitle

\begin{abstract}

\normalfont\itshape The Internet of Things (IoT) promotes the communication
among heterogeneous entities, from small sensors to Cloud systems. However,
this is realized using a wide range of communication media and data protocols,
usually incompatible with each other. Thus, IoT systems tend to grow as
homogeneous isolated platforms, which hardly interact. To achieve a higher
degree of interoperability among disparate IoT platforms, we propose a
language-based approach for communication technology integration. We build on
the Jolie programming language, which allows programmers to easily make the
same logic work over disparate communication stacks in a declarative, dynamic
way. Jolie currently supports the main technologies from Service-Oriented
Computing, such as TCP/IP, Bluetooth, and RMI at transport level, and HTTP and
SOAP at application level. As technical result, we integrate in Jolie the two
most adopted protocols for IoT communication, i.e., CoAP and MQTT. In this
paper, we report our experience and we present high-level concepts valuable
both for the general implementation of interoperable systems and for the
development of other language-based solutions.
\end{abstract}

\section{Introduction}
\label{sec:intro}
\input{intro}

\section{Jolie for IoT}
\label{sec:challenges}
\input{challenges}

\section{Supporting CoAP in Jolie}
\label{sub:coap}
\input{coap}

\section{Supporting MQTT in Jolie}
\label{sub:mqtt}
\input{mqtt}

\section{Related Work}
\label{sec:related}
\input{related}

\vspace{1em}
\section{Discussion and Conclusion}
\label{conclusion}
\input{conclusion}

% \printbibliography
% \bibliographystyle{ieeetr}
% {\raggedright\bibliography{biblio}}

\printbibliography

\end{document}
